### 2019-10-22 ms/ms.tex
Track stats; intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Bibliography%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[style=numeric-comp,sorting=none,uniquelist=minyear,uniquename=false,maxcitenames=2,maxbibnames=5,doi=false,url=false,isbn=false,defernumbers=true,giveninits=true]{biblatex}
%Italicise et al.
\renewbibmacro*{name:andothers}{
% Based on name:andothers from biblatex.def
\ifboolexpr{ test{\ifnumequal{\value{listcount}}{\value{liststop}}} and test \ifmorenames } {\ifnumgreater{\value{liststop}}{1} {\finalandcomma}
{}%
\andothersdelim\bibstring[\emph]{andothers}} {}}
%Make 'and' an ampersand
\renewcommand*{\finalnamedelim}{%
\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}%
\addspace\&\space}
%Remove 'in:'
\renewbibmacro{in:}{}
%Remove pp. on pages numbers
\DeclareFieldFormat[article]{pages}{#1}%
%Remove unwanted entries
\AtEveryBibitem{%
  \clearfield{day}%
  \clearfield{month}%
  \clearfield{endday}%
  \clearfield{endmonth}%
  \clearfield{url}%
  \clearfield{URL}%
  \clearfield{eprint}%
}
%Remove spaces between initials
%Remove double quotes around titles
\DeclareFieldFormat*{citetitle}{#1}
\DeclareFieldFormat*{title}{#1}
\addbibresource{/home/will/Papers/Library.bib}
\begin{document}
\section*{Title page}
\textbf{Article title}: Placing species' climate envelope within its
historic context reveals remarkable idiosyncrasy of distributional
change

@@

\clearpage
\textbf{Abstract. Here we
  show ... Abstract.}

% Species have climatic tolerances that are conserved
% * Caused by deep-time evolutionary processes we can detect
% * Tested by local-scale ecological assembly
% * Shown by niche-models
Species' distributions are determined by interacting factors such as
traits, climate, land-use, biotic interactions, and dispersal
limitation \supercite{Pulliam2000,Ricklefs2004,Vellend2010}. Unpicking
the relative contributions and mechanisms of these processes is
challenging, but critical if we are to make predictions about
biodiversity as these factors change simultaneously in the
Anthropocene \supercite{Heller2009,Oliver2014,Urban2015}. While
macroevolutionary biologists frequently argue that species'
distributions tend to shown phylogenetic conservatism, the tendency
for closely-related species to have similar niches
\supercite{Wiens2004,Wiens2010}. Likewise, ecologists have shown that
species' distributions are strongly driven by climate, either through
correlative studies \supercite{xxx} or experimental studies
\supercite{xxx}. While environment and climate are clearly not the
only determinants of species' distributions, decades of research into
niche distribution models have upheld the view that they are important
\supercite{xxx}.

% Species are responding to climate change
% * Species' present-day distribution still show phylogenetic signal
% * Species show local-scale assembly changes in manipulations and long-term studies
% * Ecological niche modelers show that distributions are changing
Species' distributions are undoubtedly responding to climate
\supercite{Parmesan2003,Parmesan2006} and land-use
\supercite{Newbold2015} changes. Species' contemporary distributions
still, however, exhibit phylogenetic signal \supercite{xxx}, implying
that these changes are continuations of past conditions. Equally,
experimental studies \supercite{xxx}, and the ever-increasing number
of observational studies of ongoing climate change \supercite{xxx},
have shown that species' distributions and local-scale ecological
assembly respond to climate. Niche modelers have shown that their
models have some predictive power over species' distributions given
past relationships with climate \supercite{xxx}. That there is broad
consensus that ranges are changing does not mean there have not been
debates over the precise implications and mechanisms of such
changes. For example, the implications of range change for local-scale
biodiversity is unclear
\supercite{Vellend2013,Dornelas2014,Gonzalez2016}, we do not know
whether species' environmental \emph{limits} or \emph{central
  tendencies} are most-strongly conserved
\supercite{deCasas2017,Zanne2018}, and there is a relative paucity of
historic data with which to create historic
baselines\supercite{Newbold2015,Meineke2018}.

% Yet we do not know the extent to which species are tracking their niches
% * We know there is idiosyncracy and that niche models do not always work well
% * Could this be because we have yet to define the null model by which they should be assembling?
% * Could this be because of a breakdown in the climate vectors of species? That heterogeneity is causing problems for species to track their climate?
% * Here we present a framework that accounts for this.
This consensus of species' past and present ranges broadly tracking
the same underlying climate optima is complicated by paleo-ecological
studies that have revealed greater variability of species' historic
ranges in previous centuries and millennia than previously thought
\supercite{Williams2007,Veloz2012,Maguire2015}. Critically, while
species' present-day distributions have changed, and they are broadly
similar to past distributions, we have yet to establish a historic
benchmark by which to assess the magnitude of that tracking. Many
species' distributions span continents, while others spans kilometers,
and the velocity of climate change is not constant across those scales
\supercite{xxx}. As climate continues to change, the underlying
associations between axes of climate have changed \supercite{xxx}: if
species are, as they must be \supercite{xxx}, ultimately limited by a
single environmental axis, what are the consequences of this change
for species' distributions? We argue there is a need for a synthetic
measure of the extent to which species have tracked their niche
through time, given the changes to which they have been exposed.

\subsection*{Results and discussion}
\textbf{A comparative index of species' niche tracking.} The efficacy
of niche models to predict species' distributions is, in part, a
measure of the extent to which species are tracking climate. Here we
present a simple metric of species' degree of track, which we define
as:

@@

Where $current_\tau$ is a given quantile ($\tau$) of a species'
climatic range in the present, $past_\tau$ the same quantile of a
species' range in the present, and $projected_\tau$ the same quantile
of a species' present-day range under past climatic conditions. Thus a
$track$ of 0 would indicate a species that has tracked changing
climate perfectly, while a $track$ of 1 would be consistent with a
species that had not moved whatsoever to track changing environmental
conditions. Values between 0 and 1 suggest moderate tracking, while
values greater than 1 or less than 0 represent a change in climate
relationship greater than if the species had not moved at
all. Uncertainty in either species' past or present ranges can be
accounted for through a bootstrap procedure, whereby the past and/or
present distributions are re-sampled (with replacement) to match the
sampling of the other time-period, and the standard deviation of the
index across those samples reported.

Our approach has two main innovations: (1) a focus on the whole of a
species' distribution and on (2) expected change. (1) We suggest that
something of a disconnect is developing among related fields of
ecology: species' range modelers typically focus on species'
limits\supercite{xxx} (\emph{i.e.}, the upper and lower quantiles)
while community ecologists focus on central tendencies\supercite{xxx}
(\emph{i.e.}, the median). While there is some evidence plants' global
distributions are more strongly driven by limits
\supercite{Zanne2018}, we see no strong \emph{a priori} reason that
species can, or must, track all quantiles equally\supercite{xxx}. Thus
our approach is to check all quantiles of a species' distribution. (2)
The denominator of our index re-scales species' change around an
expectation under distributional stasis. We suggest that species'
changes should be measured on the basis of the magnitude of change to
which they are exposed. This change reflects the magnitude of
ecological and evolutionary selection, and while we suggest there are
many ways of addressing this, we suggest ours is the simplest and so
perhaps a fair starting point.

\textbf{Niche tracking across the tree of life throughout the past
  sixty years.} To assess the extent to which species have tracked
climate, we used the most comprehensively sampled time-series data for
species' distributions (GBIF\supercite{xxx}) and climate (UEA's
CRU\supercite{xxx}). We focus on only five quantiles within the data:
the \nth{5}, \nth{25}, \nth{50} (median), \nth{75}, and \nth{95}. In
the methods we give more detail of our reproducible analysis pipeline,
and in the supplementary materials we repeat our analyses with more,
and less, stringent data curation and cleaning. Here, however, we
present species' range data collected from 1955--2015 for XXX species
(XXX plants, XXX fungi, XXX mammals, XXX birds, XXX reptiles, XXX
amphibians, and XXX insects), each observation matched to one of nine
climate variables in the year in which they were sampled (mean,
minimum, maximum temperature, diurnal temperature range,
precipitation, wet-day frequency, vapor pressure, and cloud
cover). We present, in the methods and supplementary materials, a more
detailed analysis of how we accounted for potential and
well-known\supercite{xxx} biases in these data, and we emphasize that
our metric was designed explicitly to account for the known increase
in sampling effort through time in GBIF data. We also highlight that
our data support known linkages between species' climatic ranges and
phylogenetic history and traits, suggesting our data reveal true
underlying signal.

Figure \ref{violin} shows the empirical index values for our focal
taxa for mean temperature. Clear patterns emerge: a large number of
species are, to some extent, tracking climate through time (XXX\%
track at least one quantile, XXX\% track at least three, and XXX\%
track all five). Of these taxa, XXX are, on average, tracking the best
(XXX\% track at least one quantile) while XXX track the worst (XXX\%
track at least one quantile).
% To discuss: these percentages go up if we use the absolutised
% version of the index, but I'm not sure that's appropriate anymore.
Results for the other eight climate variables are comparable, and are
presented in the supplementary materials. Notably, however, many taxa
are not tracking climate through time, and many have experienced more
change than if their distributions had remained static (XXX\% have
`overshot' in at least three quantiles). These results are consistent
with null simulations we performed of species' ranges that included
some degree of climate tracking and empirically observed amounts of
sampling effort changes through time and space in the GBIF data.

@@

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{../figures/violin-tmp-50.pdf}
  \end{center}
  \caption{\doublespacing \textbf{Empirical distributions of track
      index for \nth{50} of mean temperature.} Need to add simulations
    to plot. Add phylo-pics?}
  \label{pca}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Bibliography%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[style=numeric-comp,sorting=none,uniquelist=minyear,uniquename=false,maxcitenames=2,maxbibnames=5,doi=false,url=false,isbn=false,defernumbers=true,giveninits=true]{biblatex}
%Italicise et al.
\renewbibmacro*{name:andothers}{
% Based on name:andothers from biblatex.def
\ifboolexpr{ test{\ifnumequal{\value{listcount}}{\value{liststop}}} and test \ifmorenames } {\ifnumgreater{\value{liststop}}{1} {\finalandcomma}
{}%
\andothersdelim\bibstring[\emph]{andothers}} {}}
%Make 'and' an ampersand
\renewcommand*{\finalnamedelim}{%
\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}%
\addspace\&\space}
%Remove 'in:'
\renewbibmacro{in:}{}
%Remove pp. on pages numbers
\DeclareFieldFormat[article]{pages}{#1}%
%Remove unwanted entries
\AtEveryBibitem{%
  \clearfield{day}%
  \clearfield{month}%
  \clearfield{endday}%
  \clearfield{endmonth}%
  \clearfield{url}%
  \clearfield{URL}%
  \clearfield{eprint}%
}
%Remove spaces between initials
%Remove double quotes around titles
\DeclareFieldFormat*{citetitle}{#1}
\DeclareFieldFormat*{title}{#1}
\addbibresource{/home/will/Papers/Library.bib}
\begin{document}
\section*{Title page}
\textbf{Article title}: The relative velocity of climate change
reveals remarkable idiosyncrasy of species' distributional shifts

@@

\clearpage
\textbf{To date, our understanding of how species have shifted in
  response to recent climate warming has been based on a few studies
  with limited number of species. Here we present a comprehensive,
  global overview of species' distributional responses to changing
  climate across a broad variety of taxa (animals, plants, and
  fungi). We characterise species' responses using a metric that
  describes the biotic velocity of climate change: how closely
  species' responses have tracked changing climate through time
  climate. In contrast to existing metrics that focus purely on space,
  we focus on species, and thus can examine drivers of interspecific
  variation. Here we show that species responses are highly
  idiosyncratic, and thus previous analyses assuming uniform responses
  may be misleading. This is in stark contrast to species' static
  present-day and past distributions, which show strong evidence of
  the imprint of evolutionary history and functional traits. Our
  analyses are a first step in exploring the vast wealth of empirical
  data on species' historic responses to recent climate change. We
  highlight that our findings, while surprising, are consistent with
  paleontological patterns observed at the onset of mass extinction
  events.}
% Is that last sentence a bit much?

The natural environment is changing rapidly. Over the last century,
habitats have become increasingly fragmented and
isolated\supercite{Haddad2015}, the climate has become warmer, and
extreme climate events more frequent\supercite{IPCC2014}. Biological
diversity---the species of plants and animals with which we cohabit
the Earth---has responded predictably. Present-day extinction rates
are estimated to be three orders of magnitude greater than background
rates \supercite{Pimm2014} and are projected to increase further over
the next several decades \supercite{MEA2005}. Current estimates
suggest that over at least one million species may be threatened with
extinction\supercite{IPBESland2019}, and that we have experienced a
50\% decline in animal diversity over the past 40 years
\supercite{McLellan2014}.

The main direct human-induced drivers that impact biodiversity now are
habitat loss \supercite{Newbold2015} and fragmentation
\supercite{Haddad2015}, but climate change is likely to become a
dominant driver in the next few
decades\supercite{Parmesan2003,Thuiller2007,Urban2015}. Projected
impacts of climate change on biodiversity have attracted much
attention, and it has been suggested that up to 37\% of species may be
committed to global extinction by 2050\supercite{Thomas2004}. In the
face of changing climate species must either shift in space, to track
favourable conditions, or time, for example, flowering and breeding
earlier, or migrating sooner. Evidence suggests they are doing both
\supercite{Parmesan2003,Parmesan2006,Menzel2006,Chen2011}, but not all
species are responding equally. Some species are expanding their
distributions, sometimes invasively \supercite{Hellmann2008}, and some
species are flowering later\supercite{Cook2012}; species may also be
simply becoming less predictable \supercite{Pearse2017phenology}.
% I feel this isn't just a token cite; we want to set up our result a
% little, I think
Developing robust predictions of ecological responses to climate
change thus requires data from diverse species, ecosystems, and
climates to account for these varying drivers. By contrast, current
projections frequently assume a simplistic or uniform response, and
have tended to overlook interspecific variation.
% I'm not sure I necessarily agree with this statement

We currently lack robust estimates of how species are shifting their
distributions in response to recent warming. The velocity of
temperature change---the rate a species would have to shift its range
to maintain constant temperature---has been calculated at 0.42 km/yr
\supercite{Loarie2009}. A meta-analysis of taxonomic groupsestimated
the average pole-ward shift in species' distributions has been
estimated as approximately 1.69 km/yr-1 \supercite{Chen2011},
suggesting that species might be capable of keeping pace with shifting
climate. Yet this estimate is derived from just a handful of studies,
and does not address the large variation in response among species.
Here we leverage the vast volume of distributional data stored in
digital data repositories\supercite{url_gbif} to provide the first
global synthesis of species distributional responses to climate
change. By contrasting historical niche dimensions with those of the
present day, we can characterise the true biotic velocity of climate
change, which represents the actual velocity with which species track
climate across space.

We show that, on average, species' historical (pre-1970) ranges have
warmed significantly, and that species have generally shifted their
distributions to ameliorate this warming. Yet this general pattern
masks a remarkably degree of variability in species responses: many
species' ranges are now even warmer than if their distibutions had
remained static. Further, we find that, while species' distributions
are strongly predicted by their evolutionary history and functional
traits, the same is not true of their degree of tracking. Thus while
species' static distributions are reasonably predictable, their
changes in distributions are not.

\clearpage
\subsection*{Results and discussion}
\textbf{A comparative index of species' niche tracking.} We propose to
measure the extent to which a species' distribution is tracking
climate change by standardising its observed change by the change it
would have experienced had it not shifted its distribution. Formally,
we define this as:

@@

Where $current_\tau$ is a given quantile ($\tau$) of a species'
climatic distribution across its present-day range, $past_\tau$ the
same quantile of a species' range in the past, and $projected_\tau$
the same quantile of a species' present-day range under past climatic
conditions. This index should be calculated separately for
temperature, pollution, or whatever environmental variable is of
interest. Thus a $track$ of 0 would indicate a species that has
tracked changing climate perfectly, while a $track$ of 1 would be
consistent with a species that had not moved whatsoever to track
changing environmental conditions. Values between 0 and 1 suggest
moderate tracking, while values greater than 1 or less than 0
represent a change in climate relationship greater than if the species
had not moved at all. Uncertainty in either species' past or present
ranges can be accounted for through a bootstrap procedure, whereby the
past and/or present distributions are re-sampled (with replacement) to
match the sampling of the other time-period, and the standard
deviation of the index across those samples reported (see methods).

Our approach has two main innovations: its focus on (1) the whole of a
species' distribution and(2) expected change. (1) We suggest that
something of a disconnect is developing among related fields of
ecology: species' range modelers typically focus on species'
limits\supercite{Sexton2009} (\emph{i.e.}, the upper and lower
quantiles) while community ecologists \supercite{Tilman2004} and
macro-evolutionary \supercite{Zanne2018} biologists focus on central
tendencies (\emph{i.e.}, the median). We therefore see no strong
\emph{a priori} reason that species can, or must, track all quantiles
equally, and so our approach is to check all quantiles of a species'
distribution. (2) The denominator of our index re-scales species'
change around an expectation under distributional stasis. We suggest
that species' changes should be measured on the basis of the magnitude
of change to which they are exposed. This change reflects the
magnitude of ecological and evolutionary selection, and while we
suggest there are many ways of addressing this, we suggest ours is the
simplest and so perhaps a fair starting point. We note that our index
could by re-scaling according to the distance that would need to be
travelled to match the degree of change (\emph{sensu}
\citeauthor{Loarie2009}\supercite{Loarie2009}), but prefer to present
here a dimensionless (and so unitless) index of tracking.
% ...something like this? I re-wrote everything to be about the
% "relative velocity of change" (which would have given us some fancy
% notation if we wanted it) but as we don't use distances or speeds I
% feel we ought not to?

\textbf{Niche tracking across the tree of life throughout the past
  sixty years.} To assess the extent to which species have tracked
climate, we used the most comprehensively sampled time-series data for
species' distributions (GBIF\supercite{url_gbif}) and climate (UEA's
CRU\supercite{Harris2014}). We focus on five quantiles: the \nth{5},
\nth{25}, \nth{50} (median), \nth{75}, and \nth{95}. In the methods we
give more detail of our reproducible analysis pipeline, and in the
supplementary materials we repeat our analyses with different data
curation and cleaning choices. Here, however, we present species'
range data collected from 1955--2015 for 10,700 species (4,879 plants,
1,517 fungi, 273 mammals, 433 birds, 147 reptiles and amphibians, and
3,451 insects), each observation matched to one of nine climate
variables in the year in which they were sampled (mean, minimum,
maximum temperature, diurnal temperature range, precipitation, wet-day
frequency, vapour pressure, and cloud cover). Our metric was designed
to account for known biases in the GBIF data \supercite{Beck2014}, and
we present in the supplementary materials additional analyses that
address and account for such biases.

Figure \ref{violin} shows the empirical index values for our focal
taxa for mean temperature. Clear patterns emerge: 99.5\% of species
are tracking some climate variable through time, and a large number of
species are, to some extent, tracking temperature through time (65\%
track at least one quantile, 43\% track at least two, and 4\% track
all five). The three temperature indices are the least-tracked of any
climate variable, while precipitation is the most-tracked (86\% of
species track at least one of its quantiles). Notably, however, many
taxa are not tracking climate through time, and many have experienced
more change than if their distributions had remained static (99.9\% of
species have `overshot' in at least three quantiles). As we show in
figure \ref{violin}, and explore further in the supplementary
materials, these results are consistent with our null simulations
where species' range change is a combination of essentially random
movement and a moderate degree of subsequent climate filtering.

@@

This consensus of species' past and present ranges broadly tracking
the same underlying climate optima is complicated by paleo-ecological
studies that have revealed greater variability of species' historic
ranges in previous centuries and millennia than previously thought
\supercite{Williams2007,Veloz2012,Maguire2015}.

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{../figures/violin-tmp-50.pdf}
  \end{center}
  \caption{\doublespacing \textbf{Empirical distributions of track
      index for \nth{50} of mean temperature.} Need to add simulations
    to plot. Add phylo-pics?}
  \label{violin}
\end{figure}
